---
title: "Regression basics"
output: html_document
header-includes: \AtBeginDocument{\let\maketitle\relax}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(openintro)
library(tidyverse)
library(kableExtra)
library(broom) # tidy function
library(pander)
```

### 2. Fundamentals of Linear Regression - Inference 

Last time, we discussed a few ways to estimate the slope coefficient in a linear regression model. This yields a point estimate. The other key ingredient for inference is to determine how much uncertainty there is in our estimate. For the case of $\hat{\beta_1}$, we want to know how much this estimate varies from sample to sample for the specified sample size. That is, we want to know the variance of the estimate:

\begin{align}
Var(\hat{\beta_1}).
\end{align}

To find what this is, let's just try to figure out the variance directly, using the analytical formula for $\hat{\beta_1}$ in simple linear regression:

\begin{align}
Var \bigg( \frac{\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^n (x_i - \bar{x})^2 }  \bigg).
\end{align}

To carry out this calculation, recall that $\text{var}(cx) = c^2 \text{var}(x) $. Also, recall the model assumption of independence across observations. After carrying out this calculation, we can show that,

\begin{equation}
Var(\hat{\beta}) = \frac{\hat{\sigma}^2}{(n-1) \hat{var}(x)}
\end{equation}

How does the variation in $\hat{\beta}$ respond to changes in sample size? How about to variation in the predictor?

Before moving on, let's write some code to make sure our calculations correspond with those generated by <code>lm</code>.

```{r}
set.seed(1)
n = 100 # number of observations
X = rnorm(n)
y = rnorm(n , mean = 2 + 3 * X, sd = 1 )

mod = lm(y ~ X)

pander(summary(mod))

```


```{r}
sig.sq = sum(mod$residuals^2)/df.residual(mod) # sigma squared est
sqrt(sig.sq/ (var(X) * (n-1) ) )
```

In this case, we are able to get not only the variation of the estimate but also the entire distribution,


This allows us to compute confidence intervals,

\begin{align}
P(-t_{\alpha/2} < T < t_{\alpha/2})
\end{align}




### 1. Fundamentals of Linear Regression - Estimation

#### Introduction

Suppose we want to model a response variable $Y$ in terms of some predictor variables, $(X_1, X_2, X_3)$. One very general form of the model would be:

\begin{equation}
Y = f(X_1, X_2, X_3) + \epsilon
\end{equation}

where $f$ is some unknown function and $\epsilon$ is the error. Usually the exact function $f$ is unknown and we have to make assumptions about it. One such assumption is that $f$ is a linear function, which implies the following model:

\begin{equation}
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2+ \beta_3 X_3 + \epsilon.
\end{equation}

In this model, $(\beta_0, \beta_1, \beta_2, \beta_3)$ are unknown parameters. Thus, the *estimation* problem involves trying to estimate this set of coefficients.

Although this model may seem restrictive and simple, it is an extremely useful tool for gathering insight about data. It also introduces fundamental concepts to more complex statistical/ machine learning methods.


#### Example

Let's do a simple example regression analysis. We will use the county dataset from the openintro package to assess the linear association between the proportion of high school graduates in a country and the proportion of the county living in poverty. 

```{r}
# this chunk requires tidyverse and kableextra packages
dat = county_complete %>% select(hs_grad_2019, poverty_2019) # select response, predictor
kable(head(dat)) %>%
  kable_styling(position = "center") # kable prints a dataframe 
```

A good first step to any analysis is to visually explore the data. For this particular case, it is hard to beat a scatterplot for visualization.

```{r}
ggplot(dat, aes(x= hs_grad_2019, y = poverty_2019)) + 
  geom_point(alpha = 0.5) + 
  theme_minimal() + 
  ggtitle("High School Graduation Rate versus Poverty - 2019 Data")
```

It seems like fitting a linear regression to this data is reasonable. To do so, we only need a simple function in R, <code>lm</code>. Let's fit the model and see what we get.

```{r}
mod = lm(poverty_2019 ~ hs_grad_2019, data = dat)
kable(tidy(mod)) %>% kable_styling(position = "center") # requires broom package
```

How can we write down this model? What is the interpretation?

Next, we will discuss where these estimates come from. 

#### Estimation Details

Let's briefly discuss [optimization](https://tutorial.math.lamar.edu/classes/calci/optimization.aspx). Suppose we have the function

\begin{equation}
f(x) = -x^2.
\end{equation}

and we want to know the value of $x$ that maximizes this function. How would we do this? 

The estimation problem in linear regression is no different. Let's consider the simple linear regression model

\begin{equation}
y_i = \beta x_i + \epsilon_i.
\end{equation}

Our estimate of $\beta$, which we will call $\hat{\beta}$, will be the value of $\beta$ that minimizes the sum of squared differences between the observed and the predicted values. That is, it is the $\beta$ that minimizes this function:

\begin{align}
f(\beta) &= \sum_i (y_i - \hat{y}_i)^2 \\
&= \sum_i (y_i - \beta x_i)^2 
\end{align}

What is this value?

#### Exercises 1

1. Among counties with greater than 95% high school graduates, which county has the highest unemployment rate?
2. Using the command <code>data()</code>, find a dataset that interests you, generate a hypothesis regarding a linear association between two variables, and assess the hypothesis by first visualizing (using a scatterplot) and then fitting a linear regression model.
3. (Optional) Repeat the optimization analysis above for the model $y_i = \beta_0 + \beta_1x_i$, taking partial derivatives and solving.


